---
title: "R4DS_Visualize"
output: html_notebook
---

# Geometric Objects

## 9.3.1 Exercises

### 1. What geom would you use to draw a line chart? A Boxplot? A Histogram? An area chart?

a)  Line chart

```{r line_chart}
library(tidyverse)
ggplot(mpg, aes(displ, hwy)) + 
  geom_smooth()
```

b)  Boxplot

```{r boxplot}
library(tidyverse)
ggplot(mpg, aes(displ)) +
  geom_boxplot()
```

c)  Histogram

```{r histogram}
library(tidyverse)
ggplot(mpg, aes(displ, fill = drv)) +
  geom_histogram()
```

d)  area chart

```{r area chart}
library(tidyverse)
ggplot(mpg, aes(displ, hwy, fill = drv)) +
  geom_area(alpha = 0.75)
```

### 2. Earlier in this chapter we used `show.legend` without explaining it:

    -   to look like the other charts, which hadn't a legend

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = FALSE)
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(aes(color = drv), show.legend = TRUE)
```

What does `show.legend = FALSE` do here? What happens if you remove it? Why do you think we used it earlier?

### 3. What does the `se` argument to `geom_smoot()` do?

```{r}
ggplot(mpg,aes(displ, hwy)) + 
  geom_smooth(aes(color = drv), se = TRUE)
ggplot(mpg, aes(displ, hwy)) + 
  geom_smooth(aes(color = drv),se = FALSE)
```

### 4. Recreate the R code necessary to generate the following graphs. Note

    that wherever a categorical variable is used in the plot, it’s `drv`.

```{r}
# Scatterplot and lines
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(se = FALSE)

## Grouped
ggplot(mpg, aes(displ, hwy, group = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)

## colored
ggplot(mpg,aes(displ, hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)

## scatter colored only
ggplot(mpg, aes(displ,hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(se = FALSE)

## scatter colored, line distinct
ggplot(mpg,
       aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)

## scatter only, size and stroke
ggplot(mpg, aes(displ, hwy, color = drv)) +
  geom_point(size = 4 , color = "white") +
  geom_point()
```

# 9.4 Facets

Facets for cyl subset

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl)
```

Facet grid drv over cyl

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_grid(drv ~ cyl)
```

Facet grid drv over cyl, free scaling

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_grid(drv ~ cyl, scales = "free")

```

## Exercises

1.  What happens if you facet on a continuous variable?

    ```{r}
    ggplot(mpg, aes(displ, hwy)) +
      geom_point() +
      facet_grid(~cty)
    ```

    The variable is faceted by it's values.

2.  What do the empty cells in the plot above with `facet_grid(drv ~ cyl)` mean? Run the following code. How do they relate to the resulting plot?

    ```{r}
    ggplot(mpg) +    geom_point(aes(x = drv, y = cyl))
    ```

    -   The empty cells are a combination of variables without entries/values.

    <!-- -->

    -   The cyl over drv scatter-plot shows the values where the variables share entries.

<!-- -->

3.  What plots does the following code make? What does `.` do?

```{r}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

-   the `.` defines the faceting axis of the variable in the formula

4.  Take the first faceted plot in this section:

```{r}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_wrap(~ cyl, nrow = 2)

ggplot(mpg, aes(displ, hwy, color = cyl)) +
  geom_point()
```

What are the advantages to using faceting instead of the color aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?

-   (small) groups are more discernible, yet the dataset can't be seen as a whole

    -   with size of datasets advantages and disadvantages will be pronounced

5.  Read [`?facet_wrap`](https://ggplot2.tidyverse.org/reference/facet_wrap.html). What does `nrow` do? What does `ncol` do? What other options control the layout of the individual panels? Why doesn't [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html) have `nrow` and `ncol` arguments?

```{r}

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl, nrow = 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl, ncol = 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  facet_wrap(~cyl, as.table = FALSE)
```

-   `ncol` and `nrow` set the number of rows or columns to be displayed

-   further layout options are `as.table` and `dir`

-   `facet_grid`s number of rows and columns are defined by the variables chosen

6.  Which of the following plots makes it easier to compare engine size (`displ`) across cars with different drive trains? What does this say about when to place a faceting variable across rows or columns?

    ```{r}
    ggplot(mpg, aes(x = displ)) +
      geom_histogram() +
      facet_grid(drv ~ .)  

    ggplot(mpg, aes(x = displ)) +
      geom_histogram() +   
      facet_grid(. ~ drv)
    ```

    -   place faceting variable across rows to answer "where is the difference"

    -   and across columns to answer "how much is the difference"

7.  Recreate the following plot using [`facet_wrap()`](https://ggplot2.tidyverse.org/reference/facet_wrap.html) instead of [`facet_grid()`](https://ggplot2.tidyverse.org/reference/facet_grid.html). How do the positions of the facet labels change?

```{r}
ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(mpg) +
  geom_point(aes(x = displ, y = hwy)) +
  facet_wrap(~drv, nrow = 3)

```

# 9.5 Statistical Transformations

## 9.5.1 Exercises

1.  What ist the default geom associated with `stat_summary()`? How could you rewrite the previous plot to use that geom function instead of the stat function?

    ```{r}
    ggplot(diamonds) + 
      stat_summary(
        aes(x = cut, y = depth),
        fun.min = min,
        fun.max = max,
        fun = median
      )

    diamonds |> 
      group_by(cut) |> 
      summarize(
        lower = min(depth),
        upper = max(depth),
        midpoint = median(depth)
      ) |> 
      ggplot(aes(cut, midpoint)) +
      geom_pointrange(aes(ymin = lower,
                          ymax = upper)
                      )
    ```

2.  What does `geom_col()` do? How is it different from `geom_bar()`?

    ```{r}
    ggplot(diamonds, aes(cut, price)) +
      geom_col()

    ggplot(diamonds, aes(cut)) +
      geom_bar(aes(weight = price))

    ```

3.  Most geoms and stats come in pairs that are almost always used in concert. Make a list of all the pairs. What do they have in common? (Hint: Read through the documentation.)

    <div>

    | Geom       | Stat       |
    |------------|------------|
    | bar        | count      |
    | bin2d      | bin2d      |
    | boxplot    | boxplot    |
    | contour    | contour    |
    | count      | sum        |
    | density    | density    |
    | density_2d | density_2d |
    | function   | function   |
    | histogram  | stack      |
    | hex        | binhex     |
    | qqline     | qqline     |
    | quantile   | quantile   |
    | smooth     | smooth     |

    </div>

4.  What variables does `stat_smooth()` compute? What arguments control its behavior?

-   function and it's confidence interval for dataset

    -   method

    -   formula

    -   span

    -   level

```{r}
?stat_smooth()
```

5.  In our proportion bar chart, we needed to set `group = 1`. Why? In other words, what is the problem with these two graphs?

```{r}
ggplot(diamonds, aes(x = cut, y = after_stat(prop))) + 
  geom_bar()

ggplot(diamonds, aes(x = cut, fill = color, y = after_stat(prop))) + 
  geom_bar()

ggplot(diamonds, aes(cut, after_stat(prop), group = 1)) + 
  geom_bar()

ggplot(diamonds, aes(cut, after_stat(prop), fill = color, group = color)) +
  geom_bar()
```

# 9.6 Position adjustments

## Barplots

```{r}
ggplot(mpg, aes(drv, fill = class)) +
  geom_bar()

ggplot(mpg, aes(drv, fill = class)) +
  geom_bar(position ="identity")

ggplot(mpg, aes(drv, fill = class)) +
  geom_bar(position ="identity", alpha = 0.25)

ggplot(mpg, aes(drv, color = class)) +
  geom_bar(position = "identity", fill = NA)

ggplot(mpg, aes(drv, fill = class)) +
  geom_bar(position ="fill")

ggplot(mpg, aes(drv, fill = class)) +
  geom_bar(position ="dodge")


```

## Scatterplot

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(position = "jitter")

ggplot(mpg, aes(displ, hwy)) +
  geom_jitter()
```

## 9.6.1 Exercises

### 1. What is the problem with the following plot? How could you improve it?

```{r}
# Points are stacked instead of precise, average less visible
ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point()

# position argument jitter for precise points
ggplot(mpg, aes(cty, hwy)) +
  geom_point(position = "jitter")
## or use geom_jitter directly
ggplot(mpg, aes(cty, hwy)) +
  geom_jitter()
```

### 2. What, if anything, is the difference between the two plots? Why?

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(position = "identity")

# argument position ="identity" is geom_point default
```

### 3. What parameters to `geom_jitter()` control the amount of jittering?

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_jitter(width = 0.25, height = 0.25)

ggplot(mpg, aes(displ, hwy)) +
  geom_jitter(width = 0.25, height = 0.25)

ggplot(mpg, aes(displ, hwy)) +
  geom_jitter(width = 0.5, height = 0.5)
```

### 4. Compare and contrast `geom_jitter()` with `geom_count()`.

```{r}
# jitter sets overlapping point apart
ggplot(mpg, aes(displ, hwy)) +
  geom_jitter()
# count displays number of overlapping points by size
ggplot(mpg, aes(displ, hwy)) +
  geom_count()

ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_jitter()

# can use jitter as part of count to set groupin apart
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_count(position = "jitter")
```

### 5. What's the default position adjustment for `geom_boxplot()`? Create a visualization of the `mpg` dataset that demonstrates it.

```{r}
# dodge2 is default
ggplot(mpg, aes(displ, hwy, group = class, fill = class)) +
  geom_boxplot()

ggplot(mpg, aes(displ, hwy, group = class, fill = class)) +
  geom_boxplot(position = "dodge2")

# dodge impractical for x values overlap
ggplot(mpg, aes(hwy, hwy, group = class, fill = class)) +
  geom_boxplot(position = "dodge")
# same result with categorical values
ggplot(mpg, aes(class, hwy, group = class, fill = class)) +
  geom_boxplot(position = "dodge")
# boxplot ng 4 discrete values
ggplot(mpg, aes(cyl, hwy, group = class, fill = class)) +
  geom_boxplot(position = "dodge2")

```

# 9.7 Coordinate systems

```{r}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()

```

```{r}
bar <- ggplot(data = diamonds) +
  geom_bar(
    mapping = aes(clarity, fill = clarity),
    show.legend = FALSE,
    width = 1
  ) +
  theme(aspect.ratio = 1)

  bar + coord_flip()
  bar + coord_polar()
```

## 9.7.1 Exercises

### 1. Turn a stacked bar chart into a pie chart using `coord_polar()`.

```{r}
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  coord_polar()
```

### 2. What's the difference between `coord_quickmap()` and `coord_map()`?

```{r}
# coord_map needs coorect projection as argument
ggplot(map_data("nz"), aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_map()
# coord_quickmap approximates the projection
ggplot(map_data("nz"), aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```

### 3. What does the following plot tell you about the relationship between city and highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?

```{r}
# hwy / cty > 1; more hwy, the more cty
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
# coord_fixed = axis-ratio 1:1
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline()
# geom_abline = baseline 1:1-Ratio
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  coord_fixed()
```

# The layered grammar of graphics

Template

    ggplot(data = <DATA>) + 
      <GEOM_FUNCTION>(
         mapping = aes(<MAPPINGS>),
         stat = <STAT>, 
         position = <POSITION>
      ) +
      <COORDINATE_FUNCTION> +
      <FACET_FUNCTION>

The grammar of graphics is based on the insight that you can uniquely describe *any* plot as a combination of a dataset, a geom, a set of mappings, a stat, a position adjustment, a coordinate system, a faceting scheme, and a theme.
